\Huge State-Space Models and Particle Filters

### 2021-08-10 prova_knit_finale.tex
brutti file

@@

\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures

\title{\Huge State-Space Models and Particle Filters}
\author{Edoardo Marcelli \and Roberto Colarieti \and Alexei
Verkhovtsev \and Michele Bolla \and Gabriele Romano \and Benedetta
Bruni}
\date{21 luglio, 2021}

@@

\mainmatter
\chapter*{Introduction}

\chapter{Markov Processes and State-Space Models}

@@

Let \((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) and
\((\mathcal{Y}, \mathcal{B}(\mathcal{Y}))\) be two (eventually equal)
measurable spaces.

@@

\section{State-Space Models}

@@

Assume now that there exists a dominating measures \(\nu(dy)\) such that
\(F_t(x_t,dy_t)=f_t(y_t\mid x_t)\nu(dy_t)\) for all \(t.\) Then by the
last equality of definition \ref{state-space-def}, the joint
distribution of \((X_{0:T},Y_{0:T})\) is
\begin{equation}\label{state-space-densities}
   \mathbb{P}_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T})=\mathbb{P}_T(dx_{0:T})\prod\limits_{t=0}^T f_t(y_t\mid x_t)\prod\limits_{t=0}^T\nu(dy_t).
\end{equation} Integrating away \(X_{0:T},\) one gets the marginal
density \begin{align*}\label{state-space-marginal-y}
  \mathbb{P}_T(dy_{0:T}) &= \mathbb E_{\mathbb{P}_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\nu(dy_s) \\
                &= p_t(y_{0:t})\prod\limits_{s=0}^t\nu(dy_s),
\end{align*}\\
where \(p_t(y_{0:t})\) is the likelihood function. Dividing the
left-hand side of equation \ref{state-space-densities} by the left-hand
side of equation \ref{state-space-marginal-y} and the right-hand side of
equation \ref{state-space-densities} by the right-hand side of equation
\ref{state-space-marginal-y}, one obtains the filtering distribution
\begin{equation}
\mathbb{P}_t(X_{t}\in dx_{t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\mathbb{P}_t(dx_{0:t}).
\end{equation}

\chapter{Linear-Gaussian case}

@@

Suppose that \(\mathcal X=\mathbb R^{d_x}\) and
\(\mathcal Y=\mathbb R^{d_y}.\) Let
\(U_t\thicksim\mathcal N(0,\Sigma_t)\) and
\(V_t\thicksim\mathcal N(0,R_t)\) be two independent white-noise errors
and assume the initial distribution
\(X_0\thicksim\mathcal N(0,\Sigma_0).\) We have a linear Gaussian model
if \begin{align}
X_t &= A_tX_{t-1}+U_t \\
Y_t &= B_tX_t + V_t.
\end{align} Note that \(X_t\mid X_{t-1}\) and \(Y_t\mid X_t\) are both
normally distributed.\\

\begin{lemma}\label{kalman-filter}
  Let $X\thicksim\mathcal N(m_0,Q_0)$ and $Y\mid X\thicksim\mathcal N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
  Then, it holds that
  \begin{align}
    X\mid Y=y&\thicksim\mathcal N(m_1,Q_1),\\
    Q_1&= Q_0 (I_{d_x} - B'(B Q_0 B'+ R)^{-1} B Q_0) \\
    m_1&=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
  \end{align}
\end{lemma}
\begin{proof*}
   The lemma can be proved using Bayes' Theorem. Suppose $X\thicksim\mathcal N(m_0, Q_0).$ It is easy to see that
\begin{equation} \label{normal-distrib-rewrite}
        p(x)\propto \text{exp}\Big(-\frac{1}{2}X'AX + X'B\Big),
\end{equation}
with $A=Q_0^{-1}$ and $B=Am_0.$
Now, by Bayes' Theorem,

    \begin{align*}
        p(x\mid Y=y)  &\propto p(x) p(y\mid x) \\
                    &\propto \text{exp}\Big(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}(y-BX)'R^{-1}(y-BX)) \\
                    &\propto \text{exp}(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}X' B' R^{-1}BX+X' B' R^{-1}y)\\
                    &=\text{exp}(-\frac{1}{2}X'(Q_0^{-1}+B' R^{-1}B)X+
                    X'(Q_0^{-1}m_0+B' R^{-1}y)\Big).
    \end{align*}

Hence, it follows that 
\begin{equation*}
    Q_1=Var(X\mid Y=y)= (Q_0^{-1}+ B' R^{-1} B)^{-1}=Q_0-Q_0 B'(B Q_0 B'+R)^{-1}B Q_0
\end{equation*}
and 
    \begin{align*}
        m_1 &=\mathbb E(X\mid Y=y) = Q_1^{-1}(Q_0^{-1}m_0+B' R^{-1}y)\\
            &= (Q_0-Q_0B'(B Q_0B'+R)^{-1}B Q_0)(Q_0^{-1}m_0+B' R^{-1}y)\\
            &= m_0-Q_0B'(B Q_0B'+R)^{-1}B m_0 + Q_0 B' R^{-1}y - Q_0B'(B Q_0B'+R)^{-1}B Q_0 B' R^{-1}y \\
            &=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
    \end{align*}
\end{proof*}

Assume that
\(\mathbb{P}_{t-1}(dx_{t-1}\mid Y_{0:t-1}=y_{0:t-1})=\mathcal N(m_{t-1},Q_{t-1}).\)
Then, using the transition equation of the latent variable, by the
linearity of the Gaussian distribution, we have that the predictive
distribution is \begin{equation}
\mathbb{P}_{t-1}(dx_{t}\mid Y_{0:t-1}=y_{0:t-1})=\mathcal N(A_tm_{t-1},A_tQ_{t-1}A_t'+\Sigma_t).
\end{equation} Letting \(E_T=A_tQ_{t-1}A_t'+\Sigma_t,\) and applying
lemma \ref{kalman-filter}, it follows that the filtering distribution is
\begin{align*}
  \mathbb{P}_{t}(X_t\in dx_{t}\mid Y_{0:t-1}&=y_{0:t-1})=\mathcal N(m_t,Q_t), \text{where} \\ Q_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
  m_t &= [I_{d_x} - E_tB_t'(B_tE_tB_t'+R_t)^{-1} B_t]A_tm_{t-1}+E_tB_t'(B_tE_tB_t'+R_t)^{-1}y_t.
\end{align*}

@@

The aim of this subsection is to provide a practical illustration of the
algorithms discussed in the main chapters. For the sake of simplicity we
use for implementations a random walk plus noise model, i.e.~the most
basic form of linear Gaussian state-space model.\\
\begin{align}
y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \label{eq:1} \\
x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \label{eq:2} \\
x_{0} & \sim N(m_{0},C_{0})
\end{align} As already mentioned before, in this case the filtering
distribution can be computed in closed form solutions using the Kalman
filter. However, this toy example will be used also to illustrate more
involved filtering strategies described in this work. We believe indeed
that it represents a useful starting point to understand the logic of
the algorithms which may be eventually replicated when dealing with more
complex models.\\
The filtering strategy is applied to 50 simulated data. Figure
\ref{fig:myfig} shows the simulated true states sequence assuming as
data generating process the Equation \eqref{eq:2} with \(\tau^{2}=1\)
and simulated observed sequence process form Equation \eqref{eq:1} with
\(\sigma^{2}=1\).

@@

\begin{algorithm} Kalman Filter for Random Walk plus Noise Model
\begin{itemize}
\item Initialize $\theta_{0} \sim N(m_{0},C_{0})$
\item For $t=1,...,n$:
\begin{enumerate}
\item Compute the one-step-ahead state predictive distribution at time $t-1$, notice that if $t=1$ no data have been observed yet and therefore $x_{1}|x_{0}\sim N(a_{1},R_{1})$ otherwise
\begin{align*}
x_{t}|y_{1:t-1} & \sim N(a_{t},R_{t})\\
a_{t} & = m_{t-1}\\
R_{t} & = C_{t-1}+\tau^2
\end{align*}
\item Compute the filtering distribution at time $t$ as $p(x_{t}|y_{1:t}) \propto p(x_{t}|y_{1:t-1})p(y_{t}|x_{t})$, i.e. the product of the one-step-ahead state predictive distribution and the likelihood
\begin{align*}
x_{t}|y_{1:t} & \sim N(m_{t},C_{t}) \\
m_{t} & = \big(1-\frac{R_{t}}{R_{t}+\sigma^2}\big)a_{t}+\frac{R_{t}}{R_{t}+\sigma^2}y_{t} \\
C_{t} & = \frac{R_{t}}{R_{t}+\sigma^2}\sigma^2
\end{align*}
\end{enumerate}
\end{itemize}
\end{algorithm}

@@

\hfill\break
\texttt{DLM(data,sig2,tau2,m0,C0)}\\

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{DLM}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,sig2,tau2,m0,C0)\{}
\NormalTok{  n  }\OtherTok{=} \FunctionTok{length}\NormalTok{(data)}
\NormalTok{  m  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,n)}
\NormalTok{  C  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,n)}
  \ControlFlowTok{for}\NormalTok{ (t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{n)\{}
    \ControlFlowTok{if}\NormalTok{ (t}\SpecialCharTok{==}\DecValTok{1}\NormalTok{)\{}
\NormalTok{      a }\OtherTok{=}\NormalTok{ m0}
\NormalTok{      R }\OtherTok{=}\NormalTok{ C0 }\SpecialCharTok{+}\NormalTok{ tau2}
\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{}
\NormalTok{      a }\OtherTok{=}\NormalTok{ m[t}\DecValTok{{-}1}\NormalTok{]}
\NormalTok{      R }\OtherTok{=}\NormalTok{ C[t}\DecValTok{{-}1}\NormalTok{] }\SpecialCharTok{+}\NormalTok{ tau2}
\NormalTok{    \}}
\NormalTok{    A }\OtherTok{=}\NormalTok{ R}\SpecialCharTok{/}\NormalTok{(R}\SpecialCharTok{+}\NormalTok{sig2)}
\NormalTok{    m[t] }\OtherTok{=}\NormalTok{ (}\DecValTok{1}\SpecialCharTok{{-}}\NormalTok{A)}\SpecialCharTok{*}\NormalTok{a }\SpecialCharTok{+}\NormalTok{ A}\SpecialCharTok{*}\NormalTok{y[t]}
\NormalTok{    C[t] }\OtherTok{=}\NormalTok{ A}\SpecialCharTok{*}\NormalTok{sig2}
\NormalTok{  \}}
  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{m=}\NormalTok{m,}\AttributeTok{C=}\NormalTok{C))}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

In Figure \ref{fig:myfig2} below filtered states estimated using Kalman
Filter with \(x_{0} \sim N(0,100)\) and \(\sigma^{2}=\tau^{2}=1\) are
compared to the true states values. Notice how closely the filtered
states follow the observations and the goodness of the approximation of
the true states. 95 percent credible intervals are computed as
\[[E(\theta_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})},E(\theta_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})}]\]

@@

\chapter{Particle Filter}

In the previous chapter we explained how the Kalman Filter is applied to
the case of a linear model with Gaussian errors to find closed-form
solution for the filtering and the predictive distributions.
Nonetheless, the majority of time-series models have a non-linear
structure and errors might follow non-Gaussian distributions.\\
In this chapter we describe \textit{particle filter} as an algorithm
that allows to update sequentially the filtering and the predictive
distributions in the non-linear and non-Gaussian case. Additionally, the
Kalman Filter can be seen as a special case of the more general particle
filter.\\
In section \ref{pf_intro} we introduce the fundamental notions of
\textit{importance sampling}, \textit{sequential importance sampling}
and \textit{resampling}. Then, in section \ref{pf_main} we describe
three particle filtering algorithms: the \textit{bootstrap}, the
\textit{guided} and the \{auxiliary particle filters\}. Instead, in
section \ref{pf_lui_west} we describe the \{Liu and West filter\}, when
particle filtering is used to estimate model's parameters. Finally, in
section \ref{pf_converg} we establish some general convergence results
for Particle Filter - Monte Carlo estimators.

@@

Let us consider two probability spaces:
\((\mathcal X,\mathcal B(\mathcal X),\mathbb Q)\) and
\((\mathcal X,\mathcal B(\mathcal X),\mathbb M)\) and let the measurable
function \(w:\mathcal X\rightarrow \mathbb R\) be proportional to the
Radom-Nikodym derivative between \(\mathbb Q\) and \(\mathbb M\), i.e.,
\(w(x)\propto\frac{\mathbb Q(dx)}{\mathbb M(dx)}\).\\
The following result represents the starting point for the idea of
importance sampling.

@@

\begin{enumerate}
    \item Sample $N$ values from the proposal measure: $X^n\thicksim \mathbb M$ for $n=1,...,N$.
    \item Compute the associated normalized weights: $W(X^n)$ for $n=1,...,N$
    \item Compute the estimator:
    \begin{equation*}
        \hat \varphi= \frac{1}{N} \sum_{n=1}^N\varphi (X^n)W(X^n).
    \end{equation*}
\end{enumerate}

@@

\begin{enumerate}
    \item Sample $N$ values from the proposal measure: $X^n\thicksim \mathbb M$ for $n=1,...,N$.
    \item Compute the associated \textit{un-normalized} weights: $w(X^n)$ for $n=1,...,N$.
    \item Normalize the weights, for $n=1,...,N$:
    \begin{equation*}
        W^n:=\frac{w(X^n)}{\sum_{m=1}^Nw(X^m)}
    \end{equation*}
    \item Compute the estimator:
    \begin{equation*}
        \hat \varphi_{AN}=\sum_{n=1}^N\varphi (X^n)W^n
    \end{equation*}
\end{enumerate}

>>>>

\usepackage{graphicx,grffile}

@@

\title{\Huge State-Space Models and Particle Filters}
\author{Edoardo Marcelli \and Roberto Colarieti \and Alexei Verkhovtsev \and Michele Bolla \and Gabriele Romano \and Benedetta Bruni}
\date{10 Agosto, 2021}

@@

\mainmatter
\tableofcontents

\chapter*{Introduction}

State-Space models represent an extremely interesting class of models
and they are used extensively, for instance, for inference and
forecasting in time-series and macroeconometric analysis. In fact, they
allow to model observational data as function of underlying states that
cannot be observed. For instance, in macroeconomic models GDP is usually
observed while the TFP sequence is the hidden state determining
output.\\
In this work we describe techniques, such as Kalman and Particle filter,
that allow to approximate probability distributions for the states from
the observed data. For example, from the sequence of observed GDP, it is
possible to ``filter'' the TFP factor.\\
Furthermore, it is possible to impose a Markov structure to State-Space
models, assuming that the states are Markov, i.e., the probability of
the current state depends on the entire sequence of past states only
through its first lag. The combination of State-Space and Markov
structures constitutes a powerful modeling assumption for inference and
forecasting. ~

Within the Monte Carlo (MC) methods, where inference is based on
repeated simulations that allow to approximate the target distributions
(needed to compute the estimator of interest), Particle Filter (PF)
Algorithms represent the most important class of procedures for
State-Space-Markov models. PF allows to sample states period-by-period,
updating sequentially the filtering distributions (i.e., the
distribution of current state conditional on all the sequence of
observations - up to the current period one -), incorporating the
information on the likelihood of the extracted states given the sequence
of observations, by means of re-weighting.\\
Moreover, the PF is a very general algorithm, since it can be applied to
non-linear models with non-Gaussian errors, while it does not allow for
a closed form solution. On the contrary, the Kalman Filter is a
filtering method that can be applied in the case of linear-Gaussian
State-Space-Markov models, and remains analytically tractable.~ In this
work we will describe both these algorithms, while our main focus
remains on Particle Filter. ~

In chapter \ref{MP_SSM_chapter} we define precisely Markov processes and
State-Space Models. In chapter \ref{KF_chapter} we describe the Kalman
Filter and in chapter \ref{PF_chapter} we analyze extensively the
Particle Filter. Finally, in chapter \ref{application_chapter}, we apply
these methods to stochastic volatility models.\\
Finally, we find it useful to integrate each section where an algorithm
is described and analyzed from a theoretical perspective with an
implementation subsection, using as an example a simple Random-Walk
State-Space model. In this way, the reader can visualize the codes and
the practical meaning of the Kalman and Particle Filter algorithms,
immediately after their theoretical description.

\chapter{Markov Processes and State-Space Models}\label{MP_SSM_chapter}

@@

Let \((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) and
\((\mathcal{Y}, \mathcal{B}(\mathcal{Y}))\) be two (eventually equal)
measurable spaces.\\

@@

\section{State-Space Models}\label{sec_state_space}

@@

Assume now that there exists a dominating measures \(\nu(dy)\) such that
\(F_t(x_t,dy_t)=f_t(y_t\mid x_t)\nu(dy_t)\) for all \(t.\) Then by the
last equality of definition \ref{state-space-def}, the joint
distribution of \((X_{0:T},Y_{0:T})\) is
\begin{equation}\label{state-space-densities}
   \mathbb{P}_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T})=\mathbb{P}_T(dx_{0:T})\prod\limits_{t=0}^T f_t(y_t\mid x_t)\prod\limits_{t=0}^T\nu(dy_t).
\end{equation} Integrating away \(X_{0:T},\) one gets the marginal
density \begin{equation}\label{state-space-marginal-y}
  \mathbb{P}_T(dy_{0:T}) = \mathbb E_{\mathbb{P}_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\nu(dy_s)
                = p_t(y_{0:t})\prod\limits_{s=0}^t\nu(dy_s),
\end{equation}\\
where \(p_t(y_{0:t})\) is the likelihood function. Dividing the
left-hand side of equation \ref{state-space-densities} by the left-hand
side of equation \ref{state-space-marginal-y} and the right-hand side of
equation \ref{state-space-densities} by the right-hand side of equation
\ref{state-space-marginal-y}, one obtains the distribution
\begin{equation}
\mathbb{P}_t(X_{0:t}\in dx_{0:t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\mathbb{P}_t(dx_{0:t}).
\end{equation}

\chapter{Linear-Gaussian case}\label{KF_chapter}

@@

Suppose that \(\mathcal X=\mathbb R^{d_x}\) and
\(\mathcal Y=\mathbb R^{d_y}.\) Let
\(U_t\thicksim\mathcal N(0,\Sigma_t)\) and
\(V_t\thicksim\mathcal N(0,\Lambda_t)\) be two independent white-noise
errors and assume the initial distribution
\(X_0\thicksim\mathcal N(0,\Sigma_0).\) We have a linear Gaussian model
if \begin{align}
X_t &= A_tX_{t-1}+U_t \\
Y_t &= B_tX_t + V_t.
\end{align} Note that \(X_t\mid X_{t-1}\) and \(Y_t\mid X_t\) are both
normally distributed.\\

\begin{lemma}\label{kalman-filter}
  Let $X\thicksim\mathcal N(m_0,C_0)$ and $Y\mid X\thicksim\mathcal N(BX,\Lambda)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
  Then, it holds that
  \begin{align}
    X\mid Y=y&\thicksim\mathcal N(m_1,C_1),\\
    C_1&= C_0 (I_{d_x} - B'(B C_0 B'+ \Lambda)^{-1} B C_0) \\
    m_1&=[I_{d_x} - C_0 B'(B C_0 B'+ \Lambda)^{-1} B ]m_0 + C_0 B'(B C_0 B'+ \Lambda)^{-1}y.
  \end{align}
\end{lemma}
\begin{proof*}
   The lemma can be proved using Bayes' Theorem. Suppose $X\thicksim\mathcal N(m_0, C_0).$ It is easy to see that
\begin{equation} \label{normal-distrib-rewrite}
        p(x)\propto \text{exp}\Big(-\frac{1}{2}x'C_0^{-1}x + x'C_0^{-1}m_0\Big).
\end{equation}
Now, by Bayes' Theorem,

    \begin{align*}
        p(x\mid Y=y)  &\propto p(x) p(y\mid x) \\
                    &\propto \text{exp}\Big(-\frac{1}{2}x' C_0^{-1}x + x' C_0^{-1}m_0-\frac{1}{2}(y-Bx)'\Lambda^{-1}(y-Bx)) \\
                    &\propto \text{exp}(-\frac{1}{2}x' C_0^{-1}x + x' C_0^{-1}m_0-\frac{1}{2}x' B' \Lambda^{-1}Bx+x' B' \Lambda^{-1}y)\\
                    &=\text{exp}(-\frac{1}{2}x'(C_0^{-1}+B' \Lambda^{-1}B)x+
                    x'(C_0^{-1}m_0+B' \Lambda^{-1}y)\Big).
    \end{align*}

Hence, it follows that 
\begin{equation*}
    C_1=Var(X\mid Y=y)= (C_0^{-1}+ B' \Lambda^{-1} B)^{-1}=C_0-C_0 B'(B C_0 B'+\Lambda)^{-1}B C_0
\end{equation*}
and 
\begin{equation*}
\begin{split}
        m_1 &=\mathbb E(X\mid Y=y) = C_1^{-1}(C_0^{-1}m_0+B' \Lambda^{-1}y)\\
            &= (C_0-C_0B'(B C_0B'+\Lambda)^{-1}B C_0)(C_0^{-1}m_0+B' \Lambda^{-1}y)\\
            &= m_0-C_0B'(B C_0B'+\Lambda)^{-1}B m_0 + C_0 B' \Lambda^{-1}y - C_0B'(B C_0B'+\Lambda)^{-1}B C_0 B' \Lambda^{-1}y \\
            &=[I_{d_x} - C_0 B'(B C_0 B'+ \Lambda)^{-1} B ]m_0 + C_0 B'(B C_0 B'+ \Lambda)^{-1}y.
\end{split}
\end{equation*}$\hfill\blacksquare$
\end{proof*}

Assume that
\(\mathbb{P}_{t-1}(dx_{t-1}\mid Y_{1:t-1}=y_{1:t-1})=\mathcal N(m_{t-1},C_{t-1}).\)
Then, using the transition equation of the latent variable, by the
linearity of the Gaussian distribution, we have that the predictive
distribution is \begin{equation}
\mathbb{P}_{t-1}(dx_{t}\mid Y_{1:t-1}=y_{1:t-1})=\mathcal N(A_tm_{t-1},A_tC_{t-1}A_t'+\Sigma_t).
\end{equation} Letting \(R_T=A_tC_{t-1}A_t'+\Sigma_t,\) and applying
lemma \ref{kalman-filter}, it follows that the filtering distribution is
\begin{align*}
  \mathbb{P}_{t}(X_t\in dx_{t}\mid Y_{1:t-1}&=y_{1:t-1})=\mathcal N(m_t,C_t), \text{ where} \\ C_t&=R_t[I_{d_x} - B_t'(B_tR_tB_t'+\Lambda_t)^{-1} B_tR_t] \\
  m_t &= [I_{d_x} - R_tB_t'(B_tR_tB_t'+\Lambda_t)^{-1} B_t]A_tm_{t-1}+R_tB_t'(B_tR_tB_t'+\Lambda_t)^{-1}y_t.
\end{align*}

In some situations, it is possible to linearize otherwise non-linear
models in order to employ this filtering method. There are, however,
cases where this is not possible or where the linearization is not
satisfactory. One such example is the stochastic volatility model. In
the following chapters, we will introduce powerful algorithms that can
be used in much more general settings.

@@

The aim of this subsection is to provide a practical illustration of the
algorithms discussed in the main chapters. For the sake of simplicity we
use for implementations a random walk plus noise model, i.e.~the most
basic form of linear Gaussian state-space model.\\
\begin{align}
y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \label{eq:1} \\
x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \label{eq:2} \\
x_{0} & \sim N(m_{0},C_{0})
\end{align} As already mentioned before, in this case the filtering
distribution can be computed in closed form solutions using the Kalman
filter. However, this toy example will be used also to illustrate more
involving filtering strategies described in this work. We believe indeed
that it represents a useful starting point to understand the logic of
the algorithms which may be eventually replicated when dealing with more
complex models.\\
The filtering strategy is applied to 50 simulated data. Figure
\ref{fig:myfig} shows the simulated true states sequence assuming as
data generating process the Equation \eqref{eq:2} with \(\tau^{2}=1\)
and simulated observed sequence process form Equation \eqref{eq:1} with
\(\sigma^{2}=1\).

@@

\begin{algorithm} Kalman Filter for Random Walk plus Noise Model
\begin{itemize}
\item Initialize $x_{0} \sim N(m_{0},C_{0})$
\item For $t=1,...,n$:
\begin{enumerate}
\item Compute the one-step-ahead state predictive distribution at time $t-1$, notice that if $t=1$ no data have been observed yet and therefore $x_{1}|x_{0}\sim N(a_{1},R_{1})$ otherwise
\begin{align*}
x_{t}|y_{1:t-1} & \sim N(a_{t},R_{t})\\
a_{t} & = m_{t-1}\\
R_{t} & = C_{t-1}+\tau^2
\end{align*}
\item Compute the filtering distribution at time $t$ as $p(x_{t}|y_{1:t}) \propto p(x_{t}|y_{1:t-1})p(y_{t}|x_{t})$, i.e. the product of the one-step-ahead state predictive distribution and the likelihood
\begin{align*}
x_{t}|y_{1:t} & \sim N(m_{t},C_{t}) \\
m_{t} & = \big(1-\frac{R_{t}}{R_{t}+\sigma^2}\big)a_{t}+\frac{R_{t}}{R_{t}+\sigma^2}y_{t} \\
C_{t} & = \frac{R_{t}}{R_{t}+\sigma^2}\sigma^2
\end{align*}
\end{enumerate}
\end{itemize}
\end{algorithm}

@@

\texttt{DLM(data,sig2,tau2,m0,C0)}\\

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{DLM<-}\ControlFlowTok{function}\NormalTok{(data,sig2,tau2,m0,C0)\{}
\NormalTok{  n  =}\StringTok{ }\KeywordTok{length}\NormalTok{(data)}
\NormalTok{  m  =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,n)}
\NormalTok{  C  =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,n)}
  \ControlFlowTok{for}\NormalTok{ (t }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\NormalTok{n)\{}
    \ControlFlowTok{if}\NormalTok{ (t}\OperatorTok{==}\DecValTok{1}\NormalTok{)\{}
\NormalTok{      a =}\StringTok{ }\NormalTok{m0}
\NormalTok{      R =}\StringTok{ }\NormalTok{C0 }\OperatorTok{+}\StringTok{ }\NormalTok{tau2}
\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{}
\NormalTok{      a =}\StringTok{ }\NormalTok{m[t}\DecValTok{-1}\NormalTok{]}
\NormalTok{      R =}\StringTok{ }\NormalTok{C[t}\DecValTok{-1}\NormalTok{] }\OperatorTok{+}\StringTok{ }\NormalTok{tau2}
\NormalTok{    \}}
\NormalTok{    A =}\StringTok{ }\NormalTok{R}\OperatorTok{/}\NormalTok{(R}\OperatorTok{+}\NormalTok{sig2)}
\NormalTok{    m[t] =}\StringTok{ }\NormalTok{(}\DecValTok{1}\OperatorTok{-}\NormalTok{A)}\OperatorTok{*}\NormalTok{a }\OperatorTok{+}\StringTok{ }\NormalTok{A}\OperatorTok{*}\NormalTok{y[t]}
\NormalTok{    C[t] =}\StringTok{ }\NormalTok{A}\OperatorTok{*}\NormalTok{sig2}
\NormalTok{  \}}
  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{m=}\NormalTok{m,}\DataTypeTok{C=}\NormalTok{C))}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

In Figure \ref{fig:myfig2} below filtered states estimated using Kalman
Filter with \(x_{0} \sim N(0,100)\) and \(\sigma^{2}=\tau^{2}=1\) are
compared to the true states values. Notice how closely the filtered
states follow the observations and the goodness of the approximation of
the true states. 95 percent credible intervals are computed as
\[[E(x_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(x_{t}|y_{1:t})},E(x_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(x_{t}|y_{1:t})}]\]

@@

\chapter{Particle Filter}\label{PF_chapter}

In the previous chapter we explained how the Kalman Filter is applied to
the case of a linear model with Gaussian errors to find closed-form
solution for the filtering and the predictive distributions.
Nonetheless, the majority of time-series models have a non-linear
structure and errors might follow non-Gaussian distributions.\\
In this chapter we describe \textit{particle filter} as an algorithm
that allows to update sequentially the filtering and the predictive
distributions in the non-linear and non-Gaussian case. Additionally, the
Kalman Filter can be seen as a special case of the more general particle
filter.\\
In section \ref{pf_intro} we introduce the fundamental notions of
\textit{importance sampling}, \textit{sequential importance sampling}
and \textit{resampling}. Then, in section \ref{pf_main} we describe
three particle filtering algorithms: the \textit{bootstrap}, the
\textit{guided} and the \textit{auxiliary particle filters}. Instead, in
section \ref{pf_liu_west} we describe the \textit{Liu and West filter},
when particle filtering is used to estimate model's parameters. Finally,
in section \ref{pf_converg} we establish some general convergence
results for Particle Filter - Monte Carlo estimators.

@@

Let us consider two probability spaces:
\((\mathcal X,\mathcal B(\mathcal X),\mathbb Q)\) and
\((\mathcal X,\mathcal B(\mathcal X),\mathbb M)\) and let the measurable
function \(w:\mathcal X\rightarrow \mathbb R\) be proportional to the
Radom-Nikodym derivative between \(\mathbb Q\) and \(\mathbb M\), i.e.,
\(w(x)\propto\frac{\mathbb Q(dx)}{\mathbb M(dx)}\).\\
The following result represents the starting point for the idea of
importance sampling. \bigskip

@@

\begin{enumerate}
    \item Sample $N$ values from the proposal measure: $X^n\thicksim \mathbb M$ for $n=1,...,N$.
    \item Compute the associated normalized weights: $W(X^n)$ for $n=1,...,N$.
    \item Compute the estimator:
    \begin{equation*}
        \hat \varphi= \frac{1}{N} \sum_{n=1}^N\varphi (X^n)W(X^n).
    \end{equation*}
\end{enumerate}

@@

\begin{enumerate}
    \item Sample $N$ values from the proposal measure: $X^n\thicksim \mathbb M$ for $n=1,...,N$.
    \item Compute the associated \textit{un-normalized} weights: $w(X^n)$ for $n=1,...,N$.
    \item Normalize the weights, for $n=1,...,N$:
    \begin{equation*}
        W^n:=\frac{w(X^n)}{\sum_{m=1}^Nw(X^m)}.
    \end{equation*}
    \item Compute the estimator:
    \begin{equation*}
        \hat \varphi_{AN}=\sum_{n=1}^N\varphi (X^n)W^n.
    \end{equation*}
\end{enumerate}

While the auto-normalized estimator is biased (since \(W^n\) involves a
ratio of random variables), it is consistent (see \ref{pf_is_conv} for
details). ~\\
It is useful to interpret IS as an algorithm that allows to find an
approximation for the probability measure \(\mathbb Q\) as a weighted
average of Dirac measures for the sampled values (\(X^n\)):
\(\delta_{X^n}(dx)\) where the weights are the auto-normalized ones
(\(W^n\)). In particular: \begin{equation}
    \mathbb Q^N(dx):=\sum_{n=1}^NW^n\delta_{X^n}(dx), ~~~X^n\thicksim \mathbb M
\end{equation} Then, the auto-normalized IS estimator
\(\hat\varphi_{AN}\) is the expected value of the function \(\varphi\)
with respect to the approximating probability measure \(\mathbb Q^N\):
\begin{equation*}
    \begin{split}
        \mathbb E_{\mathbb Q^N}(\varphi)=&\int_\mathcal X\varphi(x)\mathbb Q^N(dx)=\int_\mathcal X\varphi(x)\sum_{n=1}^NW^n\delta_{X^n}(dx)=\\
       = &\sum_{n=1}^N\int_\mathcal X\varphi(x)W^n\delta_{X^n}(dx)=\sum_{n=1}^N\varphi (X^n)W^n=\hat \varphi_{AN}
    \end{split}
\end{equation*}
